Modal Abstractions for Virtualizing Memory Addresses

### 2023-04-10 background.tex
lots of comments

\begin{figure}
    \includegraphics[width=\columnwidth]{pagetables.png}
    \caption{x86-64 page table lookups.}
    \label{fig:pagetables}
\end{figure}

On the x86-64 architecture, the input to the \textsc{MMU}'s address translation is a sparse hierarchical set of tables calle d \emph{page tables} (referring to pages of memory). As Figure \ref{fig:pagetables}\mytodo{cite AMD ref} shows, address translation proceeds by repeatedly taking designated slices of the virtual address and indexing into a table. On x86-64, standard configurations use 4 levels of page tables, labelled levels 4 through 1, with lookups in the level 1 page table resulting in the actual page of physical memory holding the requested data, and the low-order 12 bits being used to index into this page.\footnote{Technically levels 1--3 have designated names originating when there were fewer levels. We omit those names to avoid confusion, and for greater consistency with the recent addition of a 5th level on the most recent processors, where levels 4 and 5 are simply known by their numbers.}  Because address translation traverses a series of tables, the translation process or algorithm is sometimes referred to as a \emph{page-table walk}. While Figure \ref{fig:pagetables} and most of our details are specific to the x86-64 architecture, ARMv8 (a.k.a.\ \texttt{aarch64}), RISC-V, and PowerPC use similar hierarchical page tables for address translation.

@@

The page tables mapping virtual addresses to physical addresses are managed by the OS. Typically each process has its on virtul-to-physical mappings represented by its own page table, which the OS registers with the CPU by storing it in a specific register (\texttt{cr3}) as part of switching to a new process. Using different mappings, which map only disjoint portions of physical memory (with some exceptions in the next section) is how the OS ensures memory isolation between processes.

@@

\subsection{Virtual Memory Managers}
\label{sec:backgroundonvmm}
Most OS kernels have a component called a \emph{Virtual Memory Manager} (VMM)\footnote{Not to be confused with Virtual Machine Monitor. In this paper we focus on non-virtualized scenarios, but hardware virtualization extensions for both x86-64 and ARM make use of an additional set of page tables translating what a \emph{guest} considers to be its (virutalized) physical memory to actual physical memory. Our contributions should offer value in this scenario as well.}
which is responsible for setting up page table mappings and for taking action when a page fault occurs. Often, a page fault indicates a bug in a program --- for example, a null pointer dereference crashes a program not because the hardware designates it an error, but because most OSes refuse to map the first 4K worth of virtual addresses, so dereferencing \texttt{NULL} results in a page fault. In such cases, the OS will terminate the program.

In other cases the OS uses page faults to implement specialized functionality and optimizations. A key example of this is (confusingly) called \emph{paging}: saving room in physical RAM by waiting until memory addresses for code are accessed before copying them to RAM (saving process startup time); or copying memory that has not recently been accessed to disk (i.e., in a \emph{swap} file or partition) and marking those page table entries invalid so a future page fault will give the OS the chance to copy the relevant data back into memory when the program tries to access it again.

@@

For one, virtual addresses may \emph{alias} when more than one is mapped to the same physical memory. This can occur when a process requests it (e.g., via \texttt{mmap} or \texttt{shm\_open} on UNIX-style kernels). It also occurs automatically in most general-purpose kernels: while for normal programs the OS uses the MMU to isolate and abstract physical memory, the kernel's own functionality is often easier to implement if the kernel can directly access memory given its physical address, for example when interacting with hardware devices. For this reason, most kernels contain an \emph{identity map} of all physical memory in a machine. In some cases this is a literal identity map, where for every virtual address $p$, interpreting $p$ as a virtual address instead will map back to $p$ as a physical address. In other cases there is an offset involved, where to access physical address $p$ the kernel can access virtual address $p+\mathit{offset}$. This identity map saves the kernel the performance cost of having to continually add and remove transient mappings simply to briefly access a specific physical memory region. But this coexists with additional virtual addresses (those used by most kernel code and data structures), in a form of virtual address aliasing. Note that virtual address aliasing is both prevalent in general-purpose kernels, and violates the core assumption of most memory models assumed by verified compilers like CompCert, which assume no virtual address aliasing.

Another complicating factor is that portions of page tables can be shared, i.e., trees of page tables can overlap. One common way this used to happen was for one subtree of page tables to map the kernel if linked in appropriately to a higher-level page table, and then to reuse that kernel mapping portion in every process's page tables. This was mostly phased out in the wake of Spectre\mytodo{Colin finish}

@@

Most modal logics use unary operators on propositions to express that a logical claim $P$ is \emph{contingently} true in certain other circumstances, such as in other times~\cite{pnueli1977} or places~\cite{gordon2019modal}. That is, for different modal operators $M$, $M(P)$ may mean that $P$ is true in another time or place (which may be a physical place like another network node~\cite{murphy,gordon2019modal} or a conceptual place like a person's mind~\cite{hintikka1962knowledge}). A unifying concept across any modality is that they behave as applicative functors, typically satisfying (directly, or as a derived law):
\[ (P\rightarrow Q) \rightarrow M(P) \rightarrow M(Q)\]
\mytodo{Ismail, I think this means we have pure intro, $\square P \mathrel{-\ast} [r](\square P)$ if I'm using the right symbol for pure assertions }
Many modalities, so-called \emph{normal} modalities also possess introduction rules of the form $P\rightarrow M(P)$, the classic example being that if $P$ is true, then $P$ is \emph{necessarily} true ($\square P$)

@@

A relatively under-explored space of modal logics is the interaction of modal and substructural logics, in particular hybrid-style modalities in substructural logics, which has seen only minimal exploration and no prior application. We develop our ideas in the Iris program logic~\cite{jung2018}, as a result exploring additional subtleties that arise where the modality itself may entail ownership of resources, as well as interactions between our hybrid-style modality and substructural rules.  For example, Iris contains a number of modalities that distribute over separating conjunction, or for which resources can freely move into the modality (e.g., $\blacktriangleright(P)\ast Q \mathrel{-\ast} \blacktriangleright(P\ast Q)$). In our setting some of these rules apply while others do not. For example, in our setting, an assertion that involves no modalities is interpreted as holding in the current (active-on-the-CPU) address space, so clearly cannot move into arbitrary other address spaces --- unless guarded by another address space modality.

>>>>

\begin{figure}
    \includegraphics[width=\columnwidth]{pagetables.png}
    \caption{x86-64 page table lookups.}
        \todo[inline]{Colin says: I recommend the cr3 caption say "Physical Base Address" rather than just
                base address.
            If we're saying in the text that we just to L4, L3, L2, L1, then the figure should do the same.
        }
    \label{fig:pagetables}
\end{figure}

On the x86-64 architecture, the input to the \textsc{MMU}'s address translation is a sparse hierarchical set of tables 
called \emph{page tables} (referring to pages of memory). As Figure \ref{fig:pagetables} (based on Figure 5-17 of the 
AMD64 architecture manual~\cite{amd64_manual_vol2}\footnote{While x86 up through its 32-bit incarnation were due to Intel,
the x86-64 architecture as a 64-bit extension to x86 was originally due to AMD. As a result, it is sometimes also referred to as the \texttt{amd64} architecture.}) shows, 
address translation proceeds by repeatedly taking designated slices of the virtual address and indexing into a table.
The final lookup in the page tables gives the base physical address of a page of physical memory, to which the low-order
bits of the accessed virtual address are added to determine the actual physical address retrieved. 
On x86-64, standard configurations use 4 levels of page tables, labelled levels 4 through 1, with lookups in the level 
1 page table resulting in the actual page of physical memory holding the requested data, and the low-order 12 bits 
being used to index into this page.\footnote{Technically levels 1--3 have designated names originating when there were 
fewer levels. We omit those names to avoid confusion for those not already steeped in the arcana of x86 history,
 and for greater consistency with the recent addition of a 5th level 
on the most recent processors, where levels 4 and 5 are simply known by their numbers. Our formalization
only deals with 4-level page tables, but is straightforwardly extensible to 5.}  
Because address translation 
traverses a series of tables, the translation process or algorithm is sometimes referred to as a \emph{page-table walk}. 
While Figure \ref{fig:pagetables} and most of our details are specific to the x86-64 architecture, ARMv8 (a.k.a.\ 
\texttt{aarch64}), RISC-V, and PowerPC use similar hierarchical page tables for address translation.

@@

The page tables mapping virtual addresses to physical addresses are managed by the OS. Typically each process has its own 
virtual-to-physical mappings represented by its own page table, which the OS registers with the CPU by storing the (page-aligned)
physical address of the root of the page table tree (in our context, the start of the L4 table) in a 
specific register (\texttt{cr3}) as part of switching to a new process. 
Using different mappings, which map only disjoint portions of physical memory (with some exceptions in the next section) 
is how the OS ensures memory isolation between processes.

@@

\subsection{Virtual Memory Managers}
\label{sec:backgroundonvmm}
Most OS kernels have a component called a \emph{Virtual Memory Manager} (VMM)\footnote{Not to be confused with Virtual 
Machine Monitor. In this paper we focus on non-hypervisor scenarios, but hardware virtualization extensions for both 
x86-64 and ARM make use of an additional set of page tables translating what a \emph{guest} considers to be its 
(virtualized) physical memory to actual physical memory. Our contributions should offer value in this scenario as well.}
which is responsible for setting up page table mappings and for taking action when a page fault occurs. Often, a page 
fault indicates a bug in a program --- for example, a null pointer dereference crashes a program not because the hardware
 designates it an error, but because most OSes refuse to map the first 4K worth of virtual addresses, so dereferencing 
\texttt{NULL} results in a page fault. In such cases, the OS will terminate the program.

In other cases the OS uses page faults to implement specialized functionality and optimizations. A key example of this is 
(confusingly) called \emph{paging}: saving room in physical RAM and avoiding unnecessary IO 
by waiting until memory addresses for code are accessed 
before copying them to RAM (in particular, saving process startup time); 
or copying memory that has not recently been accessed to disk 
(i.e., in a \emph{swap} file or partition) and marking those page table entries invalid so a future page fault will give 
the OS the chance to copy the relevant data back into memory when the program tries to access it again.

@@

\todo[inline]{Colin says: we should chat about the paragraph below. We don't want to advertise too many
challenges of VMM verification that we don't actually solve (or rather, we'd push them to the and of the paper
as future work). But some of these I think we could split the difference with, for example by defining
a variant of virtual-pointsto that allowed virtual address aliasing, but not actually verifying any code examples
with it (just showing it's compatible with our approach).
}
For one, virtual addresses may \emph{alias} when more than one is mapped to the same physical memory. 
This can occur when a process requests it (e.g., via \texttt{mmap} or \texttt{shm\_open} on UNIX-style kernels). 
It also occurs automatically in most general-purpose kernels: while for normal programs the OS uses the MMU to isolate 
and abstract physical memory, the kernel's own functionality is often easier to implement if the kernel can directly 
access memory given its physical address, for example when interacting with hardware devices. For this reason, most 
kernels contain an \emph{identity map} of all physical memory in a machine. 
\todo[inline]{We should try defining an identity map on a whiteboard and see what breaks. If it's feasible we should
tweak and get a definition in the paper. If not, we should cut the identity map discussion.}
In some cases this is a literal identity map, 
where for every virtual address $p$, interpreting $p$ as a virtual address instead will map back to $p$ as a physical 
address. In other cases there is an offset involved, where to access physical address $p$ the kernel can access virtual
 address $p+\mathit{offset}$. This identity map saves the kernel the performance cost of having to continually add and
 remove transient mappings simply to briefly access a specific physical memory region. But this coexists with additional
 virtual addresses (those used by most kernel code and data structures), in a form of virtual address aliasing. Note 
that virtual address aliasing is both prevalent in general-purpose kernels, and violates the core assumption of most 
memory models assumed by verified compilers like CompCert, which assume no virtual address aliasing.

\todo[inline]{Colin says: Again, let's chat about the next paragraph. This might be another where we can
split the difference on extension by defining something and proving a $\ast-\ast$ kind of result for sharing
page table fragments, rather than actually proving any new code. I'm less sure about this one, though.
}
Another complicating factor is that portions of page tables can be shared, i.e., trees of page tables can overlap and
become forests. 
One common way this used to happen was for one subtree of page tables to map the kernel if linked in appropriately to 
a higher-level page table, and then to reuse that kernel mapping portion in every process's page tables. This was mostly 
phased out in the wake of Spectre, but remains relevant on correctly-implemented hardware.

@@

Most modal logics use unary operators on propositions to express that a logical claim $P$ is \emph{contingently} true 
in certain other circumstances, such as in other times~\cite{pnueli1977temporal} or places~\cite{gordon2019modal}. That is, for 
different modal operators $M$, $M(P)$ may mean that $P$ is true in another time or place (which may be a physical place 
like another network node~\cite{murphy2008type,gordon2019modal} or a conceptual place like a person's 
mind~\cite{hintikka1962knowledge}). A unifying concept across any modality is that they behave as applicative functors, 
typically satisfying (directly, or as a derived law):
\[ (P\rightarrow Q) \rightarrow M(P) \rightarrow M(Q)\]
\todo[inline]{Ismail, I think this means we have pure intro, $\square P \mathrel{-\ast} [r](\square P)$ if I'm using the right symbol for pure assertions }
\todo[inline,color=red]{The above todo comment isn't quite right.
Purity is about being able to duplicate. You had defined another typeclass/property-of-assertions that meant it didn't
care what address space it was in (like physical pointstos). That's orthogonal to purity.
}
Many modalities, so-called \emph{normal} modalities also possess introduction rules of the form $P\rightarrow M(P)$, 
the classic example being that if $P$ is true, then $P$ is \emph{necessarily} true ($\square P$).

@@

A relatively under-explored space of modal logics is the interaction of modal and substructural logics, 
in particular hybrid-style modalities in substructural logics, which has seen only minimal exploration~\todo{copy refs from fs paper} and no prior 
application. We develop our ideas in the Iris program logic~\cite{jung2018iris}, as a result exploring additional subtleties 
that arise where the modality itself may entail ownership of resources, as well as interactions between our hybrid-style 
modality and substructural rules.  For example, Iris contains a number of modalities that distribute over separating 
conjunction, or for which resources can freely move into the modality 
(e.g., $\blacktriangleright(P)\ast Q \wand \blacktriangleright(P\ast Q)$). In our setting some of these rules 
apply while others do not. For example, in our setting, an assertion that involves no modalities is interpreted as 
holding in the current (active-on-the-CPU) address space, so clearly cannot move into arbitrary other address spaces 
--- unless guarded by another address space modality.

### 2023-04-10 cpp_paper.tex
+
removed the old stuff
replace the figure
lots of comments

%%
%% This is file `sample-acmsmall-conf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall-conf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall-conf.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall]{acmart}

@@

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}

@@

\usepackage[utf8]{inputenc}
\usepackage{todonotes}
\usepackage{blindtext}
\usepackage{listings}
\lstset{
  columns=fullflexible,
  numbers=left,
  basicstyle=\ttfamily,
  keywordstyle=\color{blue}\bfseries,
  morekeywords={mov,add,call},
  emph={rsp,rdx,rax,rbx,rbp,rsi,rdi,rcx,r8,r9,r10,r11,r12,r13,r14,r15},
  emphstyle=\color{green},
  emph={[2]cr3},
  emphstyle={[2]\color{violet}},
  morecomment=[l]{;;}
}
% Hack for Colin to edit in low light: https://tex.stackexchange.com/questions/40495/invert-background-and-text-colours-across-whole-document-with-pdflatex
%Commands definitions
\newcommand{\setbackgroundcolour}{\pagecolor[rgb]{0.19,0.19,0.19}}  
\newcommand{\settextcolour}{\color[rgb]{0.77,0.77,0.77}}    
\newcommand{\invertbackgroundtext}{\setbackgroundcolour\settextcolour}
%Command execution. 
%If this line is commented, then the appearance remains as usual.
%\invertbackgroundtext

% Packages.
\usepackage{amsmath}
% If we need multiple references to a footnote, via \cref:
\usepackage{cleveref}
\crefformat{footnote}{#2\footnotemark[#1]#3}
\usepackage[T1]{fontenc}
\usepackage{fontawesome}
\usepackage[utf8]{inputenc}
\usepackage{iris}
\usepackage{mathpartir}
\renewcommand{\TirNameStyle}[1]{\hypertarget{#1}{\textsc{#1}}}
\newcommand{\RULE}[1]{\hyperlink{#1}{\textsc{#1}}\xspace}
\usepackage{xspace}
\input{macros}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Modalities for Virtualizing Memory Addresses}
\subtitle{Understanding the Locality with Modals}
%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{X X}
\authornote{Both authors contributed equally to this research.}
\email{trovato@corporation.com}
\orcid{1234-5678-9012}
\author{G.K.M. Tobin}
\authornotemark[1]
\email{webmaster@marysville-ohio.com}
\affiliation{%
  \institution{Institute for Clarity in Documentation}
  \streetaddress{P.O. Box 1212}
  \city{Dublin}
  \state{Ohio}
  \country{USA}
  \postcode{43017-6221}
}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Operating system kernels employs virtual memory subsystem which employs memory management units on each CPU to virtualize the addresses of memory regions:a logical (virtual) address which is translated to a physical address in memory by the MMU based on
kernel-controlled page tables -- a hardware-defined sparse tree-map structure -- stored in memory, which itself is accessed by the kernel through virtual addresses.
With this core mechanism, operating system kernels generalizes its memory-resource service to different contexts, mainly through the manipulating and aliasing 
same physical memory to multiple virtual addresses through the tree structure: subtrees or data pages may be shared across processes (e.g., the kernel mapping),
accessing an unmapped address produces a fault which the OS can use to kill the process or take corrective action (like add a mapping), and resume the interrupted code which constitutes the kernel's capability to restrict which memory is accessible to different processes, hide memory limits from user programs, ensure process isolation, implement paging and copy-on-write behaviors for performance
and resource controls.

So, verifying a full-featured version of the essential virtual memory manager systems is a crucial step towards achieving the progress towards handling fully verified real kernels written using these designs. 
Regarding this effort, in this paper, we introduce a modal abstraction to describe
the truth of assertions in the virtual resources corresponding to virtual address spaces: [r]P indicating that P holds in the
virtual address space rooted at r, which implicitly treats the memory state as a modal frame and the choice of
page table root as a world in Kripke-style semantics. Moreover, we define virtual points-to assertions defined in terms of physical points-to assertions
and hardware-specific address translation assertions, and paged points-to assertions. Finally, we experiment our abstractions in specifying and proving demand paging of a fresh page of zeros and context
switches between address spaces. All definitions and theorems mentioned in this paper including the operational model of a RISC-like fragment of x86-64, a simple language run on this operational model, and a logic as an instantiation of the Iris framework are mechanized inside Coq.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Computer systems organization~Embedded systems}
\ccsdesc[300]{Computer systems organization~Redundancy}
\ccsdesc{Computer systems organization~Robotics}
\ccsdesc[100]{Networks~Network reliability}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{program verification}

@@

\input{intro.tex}
%\input{motivation.tex}
\input{background.tex}
\input{syntax.tex}
\input{semantics.tex}
\input{programlogic.tex}
\input{soundness.tex}
\input{examples.tex}
\input{discussion.tex}
\input{relwork.tex}
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{vmm}
\begin{appendix}
\section{Complete Syntax}
\label{appendix:syntax}
\begin{figure}[!ht]
\newcommand{\commentary}[1]{ & \text{\small\it #1} \\}
\[
  \begin{array}{r@{\;}c@{\;}l}
    \loc & \in & \Loc \\
    \reg & \in & \regset \\
    \regval & \in & \regvaltype \\
    \val & ::= & \vunit

\\
    \instrs & ::= &
    \begin{array}[t]{@{}l@{\hspace{10mm}}l@{}}
    \begin{array}[t]{@{}ll@{}}
      \iskip
                   \commentary{no-op}
      \iseq\instr\instrs
                   \commentary{sequencing}
      \ising\instr
                   \commentary{executing}             
    \end{array}
    %&
    %\begin{array}[t]{@{}ll@{}}
     % \ialloc\lval\allocsize
      %             \commentary{heap allocation}
    %\end{array}
    \end{array}
    \\

    \ectx & ::= &
      \hole \mid
      \iseq\ectx\instrs 
    \\
  \end{array}
\]
\caption{Syntax}
\Description{Syntax}
\label{fig:syntax}
\end{figure}
\section{Operational Semantics of Selected Mov Instructions}
\label{appendix:movops}
\input{fig-semantics}

\end{appendix}

>>>>

%%
%% This is file `sample-acmsmall-conf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall-conf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall-conf.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall,screen,review,anonymous]{acmart}

@@

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{none}
%\setcopyright{acmcopyright}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}

@@

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}
\usepackage{lipsum}
\usepackage[utf8]{inputenc}
\usepackage{todonotes}
\usepackage{blindtext}
\usepackage{listings}
\lstset{
  columns=fullflexible,
  numbers=left,
  basicstyle=\ttfamily,
  keywordstyle=\color{blue}\bfseries,
  morekeywords={mov,add,call},
  emph={rsp,rdx,rax,rbx,rbp,rsi,rdi,rcx,r8,r9,r10,r11,r12,r13,r14,r15},
  emphstyle=\color{green},
  emph={[2]cr3},
  emphstyle={[2]\color{violet}},
  morecomment=[l]{;;}
}
% Hack for Colin to edit in low light: https://tex.stackexchange.com/questions/40495/invert-background-and-text-colours-across-whole-document-with-pdflatex
%Commands definitions
\newcommand{\setbackgroundcolour}{\pagecolor[rgb]{0.19,0.19,0.19}}  
\newcommand{\settextcolour}{\color[rgb]{0.77,0.77,0.77}}    
\newcommand{\invertbackgroundtext}{\setbackgroundcolour\settextcolour}
%Command execution. 
%If this line is commented, then the appearance remains as usual.
%\invertbackgroundtext
\usepackage{wrapfig}
\newtheorem*{remark}{Remark}

% Packages.
\usepackage{amsmath}
% swiped from iris.sty in the iris repo
\newcommand{\wand}{\mathrel{-\!\!\ast}}
\newcommand{\wandIff}{\mathrel{\ast\!\!{-}\!\!\ast}}
% If we need multiple references to a footnote, via \cref:
\usepackage{cleveref}
\crefformat{footnote}{#2\footnotemark[#1]#3}
\usepackage[T1]{fontenc}
\usepackage{fontawesome}
\usepackage[utf8]{inputenc}
\usepackage{iris}
\usepackage{mathpartir}
\renewcommand{\TirNameStyle}[1]{\hypertarget{#1}{\textsc{#1}}}
\newcommand{\RULE}[1]{\hyperlink{#1}{\textsc{#1}}\xspace}
\usepackage{xspace}
\input{macros}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Modal Abstractions for Virtualizing Memory Addresses}
%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{X X}
\authornote{Both authors contributed equally to this research.}
\email{trovato@corporation.com}
\orcid{1234-5678-9012}
\author{G.K.M. Tobin}
\authornotemark[1]
\email{webmaster@marysville-ohio.com}
\affiliation{%
  \institution{Institute for Clarity in Documentation}
  \streetaddress{P.O. Box 1212}
  \city{Dublin}
  \state{Ohio}
  \country{USA}
  \postcode{43017-6221}
}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Operating system kernels employ virtual memory subsystems, which use a CPU's memory management units (MMUs) to virtualize the addresses of memory regions:
a logical (virtual) address is translated to a physical address in memory by the MMU based on
kernel-controlled page tables -- a hardware-defined sparse tree-map structure -- stored in memory, which itself is 
accessed by the kernel through virtual addresses.
Operating systems manipulate these virtualized memory mappings to isolate untrusted processes,
 restrict which memory is accessible to different processes, 
hide memory limits from user programs, 
ensure process isolation, implement demand-paging and copy-on-write behaviors for performance
and resource controls.
At the same time, misuse of MMU hardware can lead to kernel crashes.

Virtual memory management (VMM) code is a critical piece of general-purpose OS kernels, but verification of this functionality
is challenging due to the complexity of the hardware interface (the page tables are updated via writes to those
memory locations, using addresses which are themselves virtualized).
Prior work on verification of VMM code has either only handled a single address space, trusted significant
pieces of assembly code, or resorted to direct reasoning over machine semantics rather than
exposing a clean logical interface.

In this paper, we introduce a modal abstraction to describe
the truth of assertions relative to a specific virtual address space: [r]P indicating that P holds in the
virtual address space rooted at r. Such modal assertions 
allow different address spaces to refer to each other, enabling complete verification of instruction sequences
manipulating multiple address spaces. Using them effectively requires working with other assertions,
such as points-to assertions in our separation logic, as relative to a given address space.
We therefore define virtual points-to relations, which mimic hardware address translation,
relative to a page table root.
We demonstrate our approach with challenging fragments of VMM code showing that our approach
handles examples beyond what prior work can address, including reasoning about
a sequence of instructions as it changes address spaces.
All definitions and theorems mentioned in this paper including the operational model of a RISC-like fragment of x86-64, 
a simple language run on this operational model, and a logic as an instantiation of the Iris framework are mechanized 
inside Coq.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
% \begin{CCSXML}
% <ccs2012>
%  <concept>
%   <concept_id>10010520.10010553.10010562</concept_id>
%   <concept_desc>Computer systems organization~Embedded systems</concept_desc>
%   <concept_significance>500</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10010520.10010575.10010755</concept_id>
%   <concept_desc>Computer systems organization~Redundancy</concept_desc>
%   <concept_significance>300</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10010520.10010553.10010554</concept_id>
%   <concept_desc>Computer systems organization~Robotics</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10003033.10003083.10003095</concept_id>
%   <concept_desc>Networks~Network reliability</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
% </ccs2012>
% \end{CCSXML}

% \ccsdesc[500]{Computer systems organization~Embedded systems}
% \ccsdesc[300]{Computer systems organization~Redundancy}
% \ccsdesc{Computer systems organization~Robotics}
% \ccsdesc[100]{Networks~Network reliability}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
% \keywords{program verification}

@@

\input{intro.tex}
%\input{motivation.tex}
\input{background.tex}
\input{syntaxsemantics.tex}
\input{programlogic.tex}
\input{soundness.tex}
\input{examples.tex}
\input{discussion.tex}
\input{relwork.tex}
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{vmm}
